Lensing in the Schwarzschild Spacetime Applicable to Neutron Stars

### 2023-09-29 src/tex/ms.tex
WIP: doing spherical rotations.

For a ray that does not approach the surface in the equatorial plane, but rather
strikes the surface at a colatitude $\theta_0$ (but still with impact parameter
$b = x \bNS$), the tangent vector $\phi$ component rotates into $\theta$ and
$\phi$ components, and the tangent is 
\begin{eqnarray}
    \label{eq:pt-final}
    p^t & \simeq & e \left( 1 + \frac{2}{\rNS} + \ldots \right) \\ 
    \label{eq:pr-final}
    p^r & \simeq & e \sqrt{1-x^2} \left( 1 + \ldots \right) \\
    \label{eq:ptheta-final}
    p^\theta & \simeq & \frac{x e \cos \theta_0}{\rNS} \left( 1 + \frac{1}{\rNS} + \ldots \right) \\
    \label{eq:pphi-final}
    p^\phi & \simeq & \frac{x e}{\rNS} \left( 1 + \frac{1}{\rNS} + \ldots \right),
\end{eqnarray}
where we have kept only terms $\order{1/\rNS}$ beyond leading order.

@@

For a neutron star that is observed to pulse with period $P$, observers at fixed
position on the surface follow coordinate trajectories in the body frame that
have 
\begin{eqnarray}
    \chi^\theta\left( \lambda \right) & = & \theta_0 \\
    \chi^r\left( \lambda \right) & = & \rNS \\
    \chi^\phi\left(\lambda \right) & = & \frac{2 \pi \chi^t\left(\lambda \right)}{P}
\end{eqnarray}
The last equation follows from the fact that $P$ is the proper time difference
between pulses arriving at infinity at fixed location on the celestial sphere;
and thus the \emph{coordinate} time difference between pulses is $\Delta t = P$.
(Beware!  This is not the \emph{proper} time between pulses as measured by
observers fixed to the neutron star surface; this latter quantity differs by a
gravitational redshift, a time dilation due to rotational motion, and a redshift
due to the radial acceleration of the rotational motion.)

The final coordinate trajectory for the fixed-position observer can be obtained
from the normalization condition:
\begin{equation}
    g_{\mu\nu} \frac{\dd \chi^\mu}{\dd \lambda} \frac{\dd \chi^\nu}{\dd \lambda} = -1,
\end{equation}
which fixes 
\begin{equation}
    \frac{\dd \chi^t}{\dd \lambda} = \frac{1}{\sqrt{1 - \frac{2}{\rNS} - \vRot^2 \sin^2 \theta_0}} = 1 + \frac{1}{2}  \vRot^2 \sin^2 \theta_0 + \frac{1 + \frac{3}{2} \vRot^2 \sin^2 \theta_0}{\rNS} + \ldots,
\end{equation}
where the last expansion is to first contributing order in $\vRot^2 = \left( 2
\pi \rNS / P \right)^2$ and $1/\rNS$.  In the expansion you can see the effects
of special relativistic time dilation, the gravitational redshift, and the
radial acceleration (the latter being proportional to $\vRot^2 / \rNS$).  Since
this expression is constant, the coordinate time of the observer increases
linearly (but not with coefficient equal to one) with proper time:
\begin{equation}
    \Delta \chi^t = \Delta \lambda \frac{\dd \chi^t}{\dd \lambda}.
\end{equation}

To convert this world-line tangent vector to the \emph{observer} coordinate
system, let the inclination of the neutron star spin be $\iota$ and the
obliquity (measered ccw from ``vertical'' in the observer frame) be $\omega$.
Then the neutron star spin three-vector has cartesian components in the observer
frame of 
\begin{equation}
    \vec{s} = \left( \cos \iota, -\sin \iota \sin \omega, \sin \iota \cos \omega \right).
\end{equation}
See the Figure \ref{fig:obs_cart}.
%%%
\begin{figure}
    \centering
        \includegraphics[width=0.9\columnwidth]{obs_cart_coord.png}
        \caption{The coordinates in the observer frame.}
        \label{fig:obs_cart}
    \end{figure}
%%%
If the incoming ray strikes the surface at a point with colatitude $\theta_0$
and longitude $\phi_0$, then the vector pointing to the impact point has
cartesian components in the observer frame of 
\begin{equation}
    \vec{r}_0 = \left( \cos \phi_0 \sin \theta_0, \sin\phi_0 \sin \theta_0, \cos \theta_0 \right).
\end{equation}
The instantaneous world-line tangent of the observer fixed to the NS surface at
the point of impact is given in cartesian coordinates by -- IT IS THE AZIMUTHAL VELOCITY? 
\begin{equation}
    \label{eq:tangent_vel}
    \vec{t}_0 = \frac{2 \pi r_{NS}}{P} \frac{\partial \chi^t}{\partial \lambda} \vec{s} \times \vec{r}_0. 
\end{equation}

@@

Plugging in the cross product into the equation \ref{eq:tangent_vel} and solving for $r-, \theta-, \phi-$ components of the vector, then, in the observer spherical coordinate system the components of the tangent vector
are given by 
\begin{eqnarray}
    \frac{\mathrm{d}\chi^t}{\mathrm{d} \lambda} & \simeq & 1 + \frac{1}{2}  \vRot^2 \sin^2 \theta_0 + \frac{1 + \frac{3}{2} \vRot^2 \sin^2 \theta_0}{\rNS} \\
    \frac{\mathrm{d} \chi^r}{\mathrm{d} \lambda} & = & 0 \\
    \frac{\mathrm{d} \chi^\theta}{\mathrm{d} \lambda} & = & \frac{\mathrm{d} \chi^t}{\mathrm{d}\lambda} \frac{2\pi}{P} \left( - \cos\iota \sin \phi_0 - \cos\phi_0 \sin \iota \sin \omega \right) \\
    \frac{\mathrm{d} \chi^\phi}{\mathrm{d} \lambda} & = & \frac{\mathrm{d} \chi^t}{\mathrm{d}\lambda} \frac{2\pi}{P} \left( -\cos\iota \cos \phi_0 \cot \theta_0  + \sin \iota \left( \cos \omega + \cot \theta_0 \sin \phi_0 \sin \omega \right) \right)
\end{eqnarray}

[I have only worked out the geometry up to here.]

>>>>

So far we have worked in the equatorial plane, with $\theta \equiv \pi/2$; but
because of the spherical symmetry of our spacetime, our results generalize.  For
a ray with two-dimensional cartesian impact parameter coordinates $x$ and $y$;
or, equivalently, with impact parameter magnitude $l = \sqrt{x^2 + y^2}$ and
angle with the positive $x$ axis $\psi = \arctan y/x$, the impact point on the
surface becomes 
\begin{eqnarray}
    \chi^\theta \left( \rNS \right) & = & \cos^{-1} \left( \sin \alpha \sin \psi \right)  \\
    \chi^\phi \left( \rNS \right) & = & \tan^{-1} \left( \cos \psi \tan \alpha \right),
\end{eqnarray}
where the angle $\alpha$ (which corresponds to $\phi$ if the orbit is in the
equatorial plane) is given in terms of the magnitude of the impact parameter
vector $l$ by
\begin{equation}
    \alpha = \sin^{-1} l + \frac{2 \left( 1 - \sqrt{1 - l^2} \right)}{l \rNS}.
\end{equation}

[Work out the rotation of the $p^\phi$ momentum vector above.]

@@

For a neutron star that is observed to pulse with period $P$, observers at fixed
position on the surface follow coordinate trajectories in the body frame that
have 
\begin{eqnarray}
    \chi^t\left( \lambda \right) & \propto & \lambda \\
    \chi^\theta\left( \lambda \right) & = & \mathrm{const} \\
    \chi^r\left( \lambda \right) & = & \rNS \\
    \chi^\phi\left(\lambda \right) & = & \frac{2 \pi \chi^t\left(\lambda \right)}{P}
\end{eqnarray}
The last equation follows from the fact that $P$ is the proper time difference
between pulses arriving at infinity at fixed location on the celestial sphere;
and thus the \emph{coordinate} time difference between pulses is $\Delta t = P$.
(Beware!  This is not the \emph{proper} time between pulses as measured by
observers fixed to the neutron star surface; this latter quantity differs by a
gravitational redshift, a time dilation due to rotational motion, and a redshift
due to the radial acceleration of the rotational motion.)

To convert this world-line tangent vector to the \emph{observer} coordinate
system, let the inclination of the neutron star spin be $\iota$ and the
obliquity (measered ccw from ``vertical'' in the observer frame) be $\omega$.
Then the neutron star spin three-vector has cartesian components in the observer
frame of 
\begin{equation}
    \vec{s} = \left( \cos \iota, -\sin \iota \sin \omega, \sin \iota \cos \omega \right).
\end{equation}
See the Figure \ref{fig:obs_cart}.
%%%
\begin{figure}
    \centering
        \includegraphics[width=0.9\columnwidth]{obs_cart_coord.png}
        \caption{The coordinates in the observer frame.}
        \label{fig:obs_cart}
    \end{figure}
%%%
If the incoming ray strikes the surface at a point with colatitude $\theta_0$
and longitude $\phi_0$, then the vector pointing to the impact point has
cartesian components in the observer frame of 
\begin{equation}
    \vec{r}_0 = \left( \cos \phi_0 \sin \theta_0, \sin\phi_0 \sin \theta_0, \cos \theta_0 \right).
\end{equation}
The instantaneous world-line tangent of the observer fixed to the NS surface at
the point of impact is given in cartesian coordinates by
\begin{equation}
    \label{eq:tangent_vel}
    \vec{t}_0 = \frac{2 \pi r_{NS}}{P} \frac{\partial \chi^t}{\partial \lambda} \vec{s} \times \vec{r}_0. 
\end{equation}

@@

Plugging in the cross product into the equation \ref{eq:tangent_vel} and solving for $r-, \theta-, \phi-$ components of the vector, then, in the observer spherical coordinate system the components of the tangent vector
are given by 
\begin{eqnarray}
    \frac{\mathrm{d} \chi^r}{\mathrm{d} \lambda} & = & 0 \\
    \frac{\mathrm{d} \chi^\theta}{\mathrm{d} \lambda} & = & \frac{\mathrm{d} \chi^t}{\mathrm{d}\lambda} \frac{2\pi}{P} \left( - \cos\iota \sin \phi_0 - \cos\phi_0 \sin \iota \sin \omega \right) \\
    \frac{\mathrm{d} \chi^\phi}{\mathrm{d} \lambda} & = & \frac{\mathrm{d} \chi^t}{\mathrm{d}\lambda} \frac{2\pi}{P} \left( -\cos\iota \cos \phi_0 \cot \theta_0  + \sin \iota \left( \cos \omega + \cot \theta_0 \sin \phi_0 \sin \omega \right) \right)
\end{eqnarray}
The time-component of the tangent vector can be obtained from the normalization condition 
\begin{equation}
    g_{\mu \nu} \frac{\mathrm{d} \chi^\mu}{\mathrm{d} \lambda} \frac{\mathrm{d} \chi^\nu}{\mathrm{d} \lambda} = -1,
\end{equation}
which gives to leading order in $1/\rNS$ and $v_\mathrm{rot} \equiv 2 \pi \rNS / P$
\begin{multline}
    \frac{d \chi^t}{\mathrm{d} \lambda} = 1 + \frac{1}{\rNS} - v_\mathrm{rot}^2  \\ 
    \times \left( -\sin ^2(\iota ) \left(\sin ^2 \theta_0 \cos ^2 \omega +\sin ^2 \omega
    \left(\cos ^2 \theta_0 \sin ^2\phi_0 +\cos ^2 \phi_0 \right)+2 \sin \theta_0 \cos \theta_0 \sin \omega \cos
    \omega \sin \phi_0 \right) \right. \\ 
    \left. +2 \sin \theta_0 \sin \iota  \cos \iota \cos \phi_0 \left(\cos \theta_0 \cos\omega- \sin\theta_0\sin\omega\sin\phi_0
    -\left(\cos ^2\iota \left(\cos^2\theta_0\cos ^2\phi_0+\sin ^2\phi_0\right)\right) \right) \right)
\end{multline}

[Edited up to here.]
